---
title: "NOAA Storm Analysis on Population Health" 
author: "David Parker"
date: "Friday, August 22, 2014"
output: html_document
keep_md: true
---

#Synopsis  

Analyzing US Storm Data to demonstrate the human health in areas where damage occurred.

#Loading and Processing the Raw Data

Using [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) from the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database, we analyze conditions that impact the health of human populations throughout the US.

###Data Processing  

###Downloading and reading a small sample of the data for column names and early classification. 
```{r, setup, echo=TRUE, cache=TRUE}
# clear the environment
rm(list=ls())
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
              temp, method = "auto", mode = "wb")
sdata <- read.table(bzfile(temp), comment.char = "#", header = TRUE, sep = ",", nrows = 10, 
                    na.strings = "", )
dim(sdata)
cnames <- names(sdata)

cnames
cnames <- sub("__", "1", cnames, fixed = TRUE)
cnames
names(sdata) <- cnames
head(sdata[1:10], 3)
head(sdata[11:20], 3)
head(sdata[21:30], 3)
head(sdata[31:37], 3)
str(sdata)
```

##Read a bigger chunk of the data  

Some of the column classes are not interpreted correctly. Read in a larger chunk of data as character classes then reclassify the columns after examination.  

```{r, echo=TRUE, cache=TRUE}
sdata <- read.table(bzfile(temp), comment.char = "#", header = TRUE, sep = ",", nrows = 1000,
                    colClasses = "character", na.strings = "", )
names(sdata) <- cnames
# Examine header with new names
head(sdata[1:10], 3)
head(sdata[11:20], 3)
head(sdata[21:30], 3)
head(sdata[31:37], 3)
# 37 cols,  build appropriate column classes 10 at a time
cclasses <- c("factor", "Date", rep("character", 2), "numeric", "character", rep("factor", 2), "numeric", "factor", 
              rep("character", 3), "numeric", "character", "numeric", "factor", "character", rep("numeric", 2),
              "factor", rep("numeric", 4), "character", "numeric", "factor",  rep("character", 2),
              rep("character", 7))
cclasses              
```

__Read entire dataset__
```{r, echo=TRUE, cache=TRUE}
library(Defaults)
# change default date format for data load
setDefaults('as.Date.character', format = '%m/%d/%Y')
sdata <- read.table(bzfile(temp), comment.char = "#", header = TRUE, sep = ",",
                    colClasses = cclasses, na.strings = "")
names(sdata) <- cnames
dim(sdata)
summary(sdata)
str(sdata)
# Data is loaded, drop the temp compressed file
unlink(temp)
````

__Make date fields__
```{r, echo=TRUE, cache=TRUE}
# compare dates for initial date concatenation
head(sdata[, c(2:4)])
sdata$BD <- 
  as.POSIXct(
    paste(
      as.POSIXct(sdata$BGN_DATE, format="%Y/%m/%d")
      , sdata$BGN_TIME)
    ,format=  "%Y-%m-%d %H:%M:%S")

sdata$BD[1:10]
# compare dates for time correction
head(sdata[, c(2:4, 38)])
# add timezone field and create clean BEGIN_DATE
sdata$BEGIN_DATE <- 
  as.POSIXct( paste(sdata$BD, sdata$TIME_ZONE) 
              , format=  "%Y-%m-%d %H:%M:%S")
# verify new date column
sdata$BEGIN_DATE[1:10]
class(sdata$BEGIN_DATE)
head(sdata[, c(2:4, 39)])
# place BEGIN_DATE in appropiate position
sdata <- sdata[, c(1, 39, 2:37)]
str(sdata)
dim(sdata)
```
