---
title: "NOAA Storm Data Analysis on Population, Property, and Crops"
author: "David Parker"
date: "Friday, August 22, 2014"
output:
  html_document:
    keep_md: yes
---

##Synopsis  

Analyzing NOAA's Storm Data to correlate types of weather events that impact human populations. The analysis will summarize data for the time period 1950 to 2011. The results will indicate hazards to people, in terms of injuries and death, as well as economic harm in terms of property and crop damage.

###Loading and Processing the Raw Data

* Using [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) extracted from the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database.  

##Data Processing   
Download the compressed data into a Temp file then read it in compressed form.  

__Read small sample__ of the data for column names and early classification:  
```{r read_sample, echo=TRUE, cache=TRUE}
temp <- tempfile()
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
              temp, method = "auto", mode = "wb")
sdata <- read.table(bzfile(temp), comment.char = "#", header = TRUE, sep = ",", nrows = 10, 
                    na.strings = "", )
dim(sdata)
cnames <- names(sdata)

cnames
cnames <- sub("__", "1", cnames, fixed = TRUE)
cnames
names(sdata) <- cnames
head(sdata[1:10], 3)
head(sdata[11:20], 3)
head(sdata[21:30], 3)
head(sdata[31:37], 3)
str(sdata)
```

__Load entire dataset__:  
```{r load_data, echo=TRUE, cache=TRUE}
library(Defaults)
# change default date format for data load
setDefaults('as.Date.character', format = '%m/%d/%Y')
# 37 cols,  build appropriate column classes 10 at a time
cclasses <- c("factor", "Date", rep("character", 2), "numeric", "character", "factor", "character", "numeric", "factor", 
              rep("character", 3), "numeric", "character", "numeric", "factor", "character", rep("numeric", 2),
              "factor", rep("numeric", 4), "character", "numeric", "character",  rep("character", 2),
              rep("character", 7))
cclasses              
# load entire dataset
sdata <- read.table(bzfile(temp), comment.char = "#", header = TRUE, sep = ",", 
                    colClasses = cclasses, na.strings = "")
names(sdata) <- cnames
dim(sdata)
summary(sdata)
# Data is loaded, drop the temp compressed file
unlink(temp)
````

__Make date field__:  
```{r make_date, echo=TRUE, cache=TRUE}
# compare dates for initial date concatenation
head(sdata[, c(2:4)])
sdata$BEGIN_DATE <- 
  as.POSIXct(
    paste(sdata$BGN_DATE, sdata$BGN_TIME, sdata$TIME_ZONE) 
    , format=  "%Y-%m-%d %H%M")
head(sdata[, c(2:4, 38)])
# place BEGIN_DATE in appropiate position
sdata <- sdata[, c(1, 38, 2:37)]
dim(sdata)
# save sdata during testing
# save(sdata, file = 'sdata.Rdata')
```

__Tidy the data__:  
```{r tidy_data, echo=TRUE, cache=TRUE}
library(plyr)
etype <- sort(unique(sdata$EVTYPE))
etype 
# clean up EVTYPE field
sdata$EVTYPE <- toupper(sdata$EVTYPE)
sdata$EVTYPE <- sub("^\\s+", "", sdata$EVTYPE)
sdata$EVTYPE <- gsub('/', ' ', sdata$EVTYPE)

# first examine health data
# summarize by EVTYPE, examine initial values > 0
health <- ddply(sdata, .(EVTYPE), summarize, SFATAL=sum(FATALITIES), SINJURY=sum(INJURIES))

health <- health[health$SFATAL > 0| health$SINJURY > 0, ]
health[order(-health$SINJURY, -health$SFATAL), ]

# Create a new Event Type field to better summarize 
# add an entry for every significant type, init w/'OTHER'
sdata$TYPE <- "OTHER"
# load primary TYPES
sdata[grepl('TORNADO', sdata$EVTYPE),  ]$TYPE <- "TORNADO"
sdata[grepl('WIND', sdata$EVTYPE),  ]$TYPE <- "WIND"
sdata[grepl('HEAT', sdata$EVTYPE),  ]$TYPE <- "HEAT"
sdata[grepl('WARM', sdata$EVTYPE),  ]$TYPE <- "HEAT"
sdata[grepl('FLOOD', sdata$EVTYPE),  ]$TYPE <- "FLOOD"
sdata[grepl('FLD', sdata$EVTYPE),  ]$TYPE <- "FLOOD"
sdata[grepl('LIGHTN', sdata$EVTYPE),  ]$TYPE <- "LIGHTNING"
sdata[grepl('ICE', sdata$EVTYPE),  ]$TYPE <- "ICE"
sdata[grepl('WINTER', sdata$EVTYPE),  ]$TYPE <- "WINTER WEATHER"
sdata[grepl('WINTRY', sdata$EVTYPE),  ]$TYPE <- "WINTER WEATHER"
sdata[grepl('FIRE', sdata$EVTYPE),  ]$TYPE <- "FIRE"
sdata[grepl('HAIL', sdata$EVTYPE),  ]$TYPE <- "HAIL"
# hurricane & typhoon is a form of tropical cyclone
sdata[grepl('HURRICANE', sdata$EVTYPE),  ]$TYPE <- "CYCLONE"
sdata[grepl('TYPHOON', sdata$EVTYPE),  ]$TYPE <- "CYCLONE"
sdata[grepl('SNOW', sdata$EVTYPE),  ]$TYPE <- "SNOW"
sdata[grepl('FOG', sdata$EVTYPE),  ]$TYPE <- "FOG"
sdata[grepl('ICY', sdata$EVTYPE),  ]$TYPE <- "FOG"
sdata[grepl('FOG', sdata$EVTYPE),  ]$TYPE <- "FOG"
sdata[grepl('BLIZZ', sdata$EVTYPE),  ]$TYPE <- "BLIZZARD"
sdata[grepl('RIP', sdata$EVTYPE),  ]$TYPE <- "RIP CURRENTS"
sdata[grepl('DUST', sdata$EVTYPE),  ]$TYPE <- "DUST STORM"
sdata[grepl('TROPIC', sdata$EVTYPE),  ]$TYPE <- "TROPICAL STORM"
sdata[grepl('RAIN', sdata$EVTYPE),  ]$TYPE <- "RAIN"
sdata[grepl('COLD', sdata$EVTYPE),  ]$TYPE <- "COLD"
sdata[grepl('HYPOTH', sdata$EVTYPE),  ]$TYPE <- "COLD"
sdata[grepl('FREEZ', sdata$EVTYPE),  ]$TYPE <- "COLD"
sdata[grepl('FROST', sdata$EVTYPE),  ]$TYPE <- "COLD"
sdata[grepl('AVALAN', sdata$EVTYPE),  ]$TYPE <- "AVALANCHE"
sdata[grepl('SEAS', sdata$EVTYPE),  ]$TYPE <- "HIGH SEAS"
sdata[grepl('SURF', sdata$EVTYPE),  ]$TYPE <- "HIGH SEAS"
# sdata[grepl('TSUN', sdata$EVTYPE),  ]$TYPE <- "TSUNAMI"
# sdata[grepl('LANDS', sdata$EVTYPE),  ]$TYPE <- "LANDSLIDE"
sdata[grepl('SURGE', sdata$EVTYPE),  ]$TYPE <- "STORM SURGE"
sdata[grepl('TSTM', sdata$EVTYPE),  ]$TYPE <- "THUNDERSTORM"
sdata[grepl('THUNDER', sdata$EVTYPE),  ]$TYPE <- "THUNDERSTORM"
sdata[grepl('WATERSP', sdata$EVTYPE),  ]$TYPE <- "WATERSPOUT"
# many DROUGHT combos, override above others in one DROUGHT category
sdata[grepl('DROUGHT', sdata$EVTYPE),  ]$TYPE <- "DROUGHT"
# See major TYPES
type <- sort(unique(sdata$TYPE))
type 
sdata <- transform(sdata, TYPE = factor(TYPE))
```

__Analyze the Tidy data__
```{r analyze_data,echo=TRUE,cache=TRUE}
# summary(sdata$BEGIN_DATE)
summary(sdata$BGN_DATE)
# summarize damage estimates for health 
health <- ddply(sdata, .(TYPE), summarize, 
                FATALITY=sum(FATALITIES, na.rm = TRUE), INJURY=sum(INJURIES, na.rm = TRUE))
health
healthy <- ddply(sdata, .((format(sdata$BGN_DATE, '%Y')), TYPE), summarize, 
                FATALITY=sum(FATALITIES, na.rm = TRUE), INJURY=sum(INJURIES, na.rm = TRUE))
colnames(healthy)[1] <- c("YEAR")
tail(healthy)
# ready for health analysis results

# now examine property & crop economic damage
# limit the data frame to only needed columns for economic data
ldata <- sdata[, c(3, 26, 27, 28, 29, 39)]
summary(ldata)
names(ldata)
edata <- ddply(ldata, .(TYPE, PROPDMGEXP, CROPDMGEXP), summarize, 
              PROPDMG = sum(PROPDMG, na.rm=T), CROPDMG = sum(CROPDMG,na.rm=T))
head(edata,3)
# examine yearly breakdown
edatay <- ddply(ldata, .(TYPE, (format(BGN_DATE, '%Y')), PROPDMGEXP, CROPDMGEXP), summarize, 
              PROPDMG = sum(PROPDMG, na.rm=T), CROPDMG = sum(CROPDMG,na.rm=T))
colnames(edatay)[2] <- c("YEAR")
tail(edatay)

# Focus on areas where damages are in Billions of dollars
bdamage <- edata[edata$PROPDMGEXP %in% ('B')|edata$CROPDMGEXP %in% ('B'), ]
# calaculate damages for crops, property and total in billions
damage <- ddply(bdamage, .(TYPE), summarize, PROPERTY = round(sum(PROPDMG)),
               CROPS = round(sum(CROPDMG)),
               DAMAGES = (round(sum(PROPDMG)) + round(sum(CROPDMG))))
# Damages in Billions
damage

# Focus on areas by Year where damages are in Billions of dollars
bdamagey <- edatay[edatay$PROPDMGEXP %in% ('B')|edatay$CROPDMGEXP %in% ('B'), ]
# calaculate damages for crops, property and total in billions
damagey <- ddply(bdamagey, .(YEAR, TYPE), summarize, PROPERTY = round(sum(PROPDMG)),
               CROPS = round(sum(CROPDMG)),
               DAMAGES = (round(sum(PROPDMG)) + round(sum(CROPDMG))))
# Damages in Billions
damagey
# ready for results
```


##Results  

```{r results_health,echo=TRUE,cache=TRUE}
library(ggplot2)
# plot summary health effects 
health2 <- health[health$FATALITY > 0 & health$INJURY > 0, ]
health2$TYPE <- factor(health$TYPE, levels=health[order(health$FATALITY), "TYPE"])

h <- ggplot(health2, aes(y = INJURY,  x = TYPE, fill = FATALITY)) + coord_flip()
h <- h + geom_bar(stat = 'identity', binwidth = 2) 
h <- h + scale_fill_gradient(low="green",high="darkgreen")
h <- h + labs(title = 'Fatalies & Injuries due to Weather Events')
h <- h + xlab('Weather Event Type (Sort on Fatality)')
h <- h + ylab('Injury Tallies (Fatalities in Red)')
h <- h + geom_text(size=3,  angle = 15, color='red', hjust=-0.5, vjust=0, aes(label=FATALITY)) 
print(h)
```   

Tornadoes are far and away the most significant threat in terms of injury and loss of life.  

```{r results_damage,echo=TRUE,cache=TRUE}
# plot summary property & crop damages
damage <- damage[order(-damage$DAMAGES), ]
# plot DAMAGES >= 1 billion 
damage2 <- damage[damage$DAMAGES >= 1, ]
damage2$TYPE <- factor(damage2$TYPE, levels=damage2[order(damage2$PROPERTY), "TYPE"])

e <- ggplot(damage2, aes(y = DAMAGES,  x = TYPE,  fill = PROPERTY)) + coord_flip()
e <- e + geom_bar(stat = 'identity', binwidth = 1) 
e <- e + scale_fill_gradient(low="green",high="darkgreen")
e <- e + ggtitle(expression(atop('Property & Crop Economic Losses due to Weather Events', 
                                 atop(italic('(Property in Red Numbers)'), ""))))
e <- e + xlab('Weather Event Type (Sort on Property)')
e <- e + ylab('Total Economic to Property & Crops Damage in Billions')
e <- e + geom_text(size=3,  angle = 15, color='red', hjust=-0.5, vjust=0, aes(label=PROPERTY)) 
print(e)
```  

Cyclones (Hurricanes and Tropical Cyclones) generate the most total losses for property and crops combined, but most of this damage is to crops. The types are sorted in terms of damage to property indicating that ice and flooding are a greater threat to property followed by cyclones.
