---
title: "NOAA Storm Analysis on Population Health" 
author: "David Parker"
date: "Friday, August 22, 2014"
output: html_document
keep_md: true
---

#Synopsis  

Analyzing US Storm Data to demonstrate the human health in areas where damage occurred.

#Loading and Processing the Raw Data

Using [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) from the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database, we analyze conditions that impact the health of human populations throughout the US.

###Data Processing  

###Downloading and reading a small sample of the data for column names and early classification. 
```{r, echo=TRUE, cache=TRUE}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
              temp, method = "auto", mode = "wb")
sdata <- read.table(bzfile(temp), comment.char = "#", header = TRUE, sep = ",", nrows = 10, 
                    na.strings = "", )
dim(sdata)
# rawnames <- readLines(bzfile(temp), 1)
cnames <- names(sdata)
cnames
cnames <- sub("__", "", cnames, fixed = TRUE)
cnames
names(sdata) <- cnames
head(sdata[1:10], 3)
head(sdata[11:20], 3)
head(sdata[21:30], 3)
head(sdata[31:37], 3)
str(sdata)
```

##Read all the data  

Some of the column classes are not interpreted correctly. Read in a larger chunk of data as character classes then reclassify the columns after examination.  

```{r, echo=TRUE, cache=TRUE}
sdata <- read.table(bzfile(temp), comment.char = "#", header = TRUE, sep = ",", nrows = 1000,
                    colClasses = "character", na.strings = "", )
names(sdata) <- cnames
unique(sdata$STATE)
unique(sdata$COUNTY)
unique(sdata$BGN_DATE)
unique(sdata$BGN_TIME)
unique(sdata$TIME_ZONE)
unique(sdata$BGN_AZI)
unique(sdata$COUNTYENDN)
unique(sdata$END_LOCATI)
unique(sdata$WFO)
unique(sdata$STATEOFFIC)
unique(sdata$ZONENAMES)
unique(sdata$PROPDMGEXP)
unique(sdata$REFNUM)
head(sdata[1:10])
head(sdata[11:20])
head(sdata[21:30])
head(sdata[31:37])
dim(sdata)
# [1] 902297     37
# 37 cols,  build classes 10 at a time
cclasses <- c("factor", rep("character", 3), "numeric", "character", rep("factor", 2), "numeric", "factor", 
              rep("character", 3), "numeric", "character", "numeric", "factor", "character", rep("numeric", 2),
              "factor", rep("numeric", 4), "character", "numeric", "factor",  rep("character", 2),
              rep("character", 7))
cclasses              
```

__Read entire dataset__
```{r, echo=TRUE, cache=TRUE}
sdata <- read.table(bzfile(temp), comment.char = "#", header = TRUE, sep = ",",
                    colClasses = cclasses, 
                    na.strings = "", )
names(sdata) <- cnames
dim(sdata)
summary(sdata)
str(sdata)
unlink(temp)
````

__Make date fields__
```{r, echo=TRUE, cache=TRUE}
# Create date from BGN_DATE with overlapping time fields & concatenate TZ
sdata$BD1 <- as.POSIXct(sdata$BGN_DATE, format="%m/%d/%Y")
sdata$BD1
head(sdata[, c(2:4, 39:40)])
sdata$BD2 <- as.POSIXct(
  paste(as.POSIXct(sdata$BD, format="%Y/%m/%d"), sdata$BGN_TIME)
  ,format="%Y-%m-%d %H%M")
sdata$BD2
head(sdata[, c(2:4, 39:41)])
sdata$BEGIN_DATE <- 
  as.POSIXct(
    paste(sdata$BD2, sdata$TIME_ZONE) )
sdata$BEGIN_DATE
head(sdata[, c(2:4, 39:42)])
class(sdata$BEGIN_DATE)             
str(sdata)
dim(sdata)
sdata$BEGIN_DATE[1:6]
```